---
title: "10_18_table_1_rep_full_data"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(readr)
library(coin)
```

## 1. check grouping
```{r, message=FALSE, warning=FALSE}
# load cleaned data
df.clean <- read_csv("UPDATED_3_trauma_data_clean.csv")
df.selected <- df.clean %>% filter(!is.na(gcs_calc_a)) 
dim(df.selected)
df.selected$updated_group <- as.factor(df.selected$updated_group)
table(df.selected$updated_group)
head(df.selected)
```

## 2. variable analysis in table 1 

```{r}
# age
df.age <- df.selected %>% group_by(updated_group) %>% summarise(age.med = median(age), age.1Q = quantile(age, 0.25), age.3Q = quantile(age, 0.75))
df.age
```

```{r}
# sex: 0=male, 1=female
df.selected$sex = ifelse(df.selected$sex=="Female",1,0)
df.sex <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.male = n()-sum(sex), perc = count.male/count)
df.sex
```

```{r}
# injury_type: 0=blunt, 1=penetrating
df.injurytype <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.penetrating = sum(injury_type), perc = count.penetrating/count)
df.injurytype
```

```{r}
# gcs
df.selected$gcs_calc_a <- as.character(df.selected$gcs_calc_a)
df.selected$gcs_calc_a[df.selected$gcs_calc_a== "<unk>"] <- NA
df.selected$gcs_calc_a <- as.numeric(df.selected$gcs_calc_a)

df.gcs <- df.selected %>% 
  group_by(updated_group) %>% 
  summarise(gcs.med = median(gcs_calc_a, na.rm = TRUE), 
            gcs.1Q = quantile(gcs_calc_a, 0.25, na.rm = TRUE), 
            gcs.3Q = quantile(gcs_calc_a, 0.75, na.rm = TRUE))
df.gcs
```

```{r}
# blood pressure
df.bp0 <- df.selected %>% 
  filter(sys_bp_a==0) %>% 
  group_by(updated_group) %>% 
  summarise(count_0=n())

df.bp <- df.selected %>% 
  group_by(updated_group) %>% 
  summarise(bp.med = median(sys_bp_a), 
            bp.1Q = quantile(sys_bp_a, 0.25), 
            bp.3Q = quantile(sys_bp_a, 0.75), 
            count = n()) %>% 
  left_join(df.bp0, by = "updated_group") %>%
  mutate(perc_0 = count_0/count)

df.bp
```

```{r}
# heart rate
df.selected$pulse_a <- as.numeric(as.character(df.selected$pulse_a))

df.hr0 <- df.selected %>% 
  filter(pulse_a==0) %>% 
  group_by(updated_group) %>% 
  summarise(count_0=n())

df.hr <- df.selected %>% 
  group_by(updated_group) %>% 
  summarise(hr.med = median(pulse_a, na.rm = TRUE), 
            hr.1Q = quantile(pulse_a, 0.25, na.rm = TRUE), 
            hr.3Q = quantile(pulse_a, 0.75, na.rm = TRUE),
            count = n())%>% 
  left_join(df.hr0, by = "updated_group") %>%
  mutate(perc_0 = count_0/count)

df.hr
```

```{r}
df.ISS <- df.selected %>% group_by(updated_group) %>% summarise(ISS.med = median(injury_serverity_score, na.rm = TRUE), ISS.1Q = quantile(injury_serverity_score, 0.25, na.rm = TRUE), ISS.3Q = quantile(injury_serverity_score, 0.75, na.rm = TRUE))
df.ISS
```

```{r}
# trauma_center_level: 0=level1, 1=level 2
df.selected$trauma_center_level <- ifelse(df.selected$trauma_center_level=="Level 1",1,0)
df.level <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.level1 = sum(trauma_center_level), perc = count.level1/count)
df.level
```

```{r}
# Mortality rate 
df.death1 <- df.selected %>% 
  filter(death=="Dead") %>% 
  group_by(updated_group) %>% 
  summarise(count_1=n())

df.death <- df.selected %>% 
  group_by(updated_group) %>% 
  summarise(count = n())%>% 
  left_join(df.death1, by = "updated_group") %>%
  mutate(perc_death = count_1/count)
df.death
```

```{r}
# Post ED destination 
# - Died in ED (6)
# - Operating room (2)
# - Interventional radiology(11)
# - Intensive care unit (1)

# Died in ED (6)
df.DE <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.DE = sum(post_ed==6), perc = count.DE/count)
df.DE

# Operating room (2)
df.OR <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.OR = sum(post_ed==2), perc = count.OR/count)
df.OR

# Interventional radiology(11)
df.IR <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.IR = sum(post_ed==11), perc = count.IR/count)
df.IR

# Intensive care unit (1)
df.ICU <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.ICU = sum(post_ed==1), perc = count.ICU/count)
df.ICU
```

```{r}
# Mortality rate at 24 hours after admission 

# eda_date was a character type, make it a date
df.selected$eda_date <- as.Date(df.selected$eda_date, format = "%m/%d/%Y")

df.selected$eda_T <- as.POSIXct(paste(df.selected$eda_date, df.selected$eda_time), format="%Y-%m-%d %H:%M:%S")
df.selected$death_T <- as.POSIXct(paste(df.selected$day_of_death, df.selected$time_of_death), format="%Y-%m-%d %H:%M:%S")

df.selected$timediff <-df.selected$death_T-df.selected$eda_T

df.24MR <- df.selected %>% group_by(updated_group) %>% summarise(count = n(), count.24MR = sum(timediff<=24*3600, na.rm = TRUE), perc = count.24MR/count)
df.24MR
```


## 4. univariate test for variables in table 1
For continuous variable, perform Kruskal–Wallis test to test whether samples in 4 groups originate from the same distribution.  
For binomial variables, preform Pearson's Chi-squared Test to test whether 4 groups have the same frequency.

#### 4.1 continuous variables

```{r}
# p-value of age
age.test <- kruskal.test(age ~ updated_group,data = df.selected)
p_val.age <- age.test$"p.value" 
ifelse(p_val.age<0.0001,"<0.0001", round(p_val.age, digits = 4))

# p-value of gcs
gcs.test <- kruskal.test(gcs_calc_a ~ updated_group,data = df.selected)
p_val.gcs <- gcs.test$"p.value"
ifelse(p_val.gcs<0.0001,"<0.0001", round(p_val.gcs, digits = 4))

# p-value of blood pressure
sbp.test <- kruskal.test(sys_bp_a ~ updated_group,data = df.selected)
p_val.sbp <- sbp.test$"p.value" 
ifelse(p_val.sbp<0.0001,"<0.0001", round(p_val.sbp, digits = 4))


# p-value of heart rate
hr.test <- kruskal.test(pulse_a ~ updated_group,data = df.selected)
p_val.hr <- hr.test$"p.value" 
ifelse(p_val.hr<0.0001,"<0.0001", round(p_val.hr, digits = 4))

# p-value of ISS
ISS.test <- kruskal.test(injury_serverity_score ~ updated_group,data = df.selected)
p_val.ISS <- ISS.test$"p.value" 
ifelse(p_val.ISS<0.0001,"<0.0001", round(p_val.ISS, digits = 4))
```

## 4.2 binary variables 

```{r, fig.height=3, fig.width=5}
df.selected$updated_group <- as.factor(df.selected$updated_group)
# p-value of sex: 0=male, 1=female
df.selected$sex <- as.factor(df.selected$sex)
test.sex <- chisq_test(sex ~ updated_group, data = df.selected)
p_val.sex <- pvalue(test.sex)
ifelse(p_val.sex<0.0001,"<0.0001", round(p_val.sex, digits = 4))

# p-value of injury type: 0=blunt, 1=penetrating
df.selected$injury_type <- as.factor(df.selected$injury_type)
test.injurytype <- chisq_test(injury_type ~ updated_group, data = df.selected)
p_val.injurytype <- pvalue(test.injurytype)
ifelse(p_val.injurytype<0.0001,"<0.0001", round(p_val.injurytype, digits = 4))

# p-value of bp_zero : 1 if sys_bp_a=0, 0 otherwise
df.selected$bp_zero <- ifelse(df.selected$sys_bp_a == 0,1,0)
df.selected$bp_zero <- as.factor(df.selected$bp_zero)
test.bp_zero <- chisq_test(bp_zero ~ updated_group, data = df.selected)
p_val.bp_zero <- pvalue(test.bp_zero)
ifelse(p_val.bp_zero<0.0001,"<0.0001", round(p_val.bp_zero, digits = 4))

# p-value of pulse_zero : 1 if pulse_a=0, 0 otherwise
df.selected$pulse_zero <- ifelse(df.selected$pulse_a == 0,1,0)
df.selected$pulse_zero <- as.factor(df.selected$pulse_zero)
test.pulse_zero <- chisq_test(pulse_zero ~ updated_group, data = df.selected)
p_val.pulse_zero <- pvalue(test.pulse_zero)
ifelse(p_val.pulse_zero<0.0001,"<0.0001", round(p_val.pulse_zero, digits = 4))

# p-value of trauma center level: 0=level1, 1=level 2
test.centerlevel <- chisq_test(factor(trauma_center_level) ~ updated_group, data = df.selected)
p_val.centerlevel <- pvalue(test.centerlevel)
ifelse(p_val.centerlevel<0.0001,"<0.0001", round(p_val.centerlevel, digits = 4))

# p-value of death: 1=dead, 0 otherwise
df.selected$death <- as.factor(df.selected$death)
test.death <- chisq_test(death ~ updated_group, data = df.selected)
p_val.death <- pvalue(test.death)
ifelse(p_val.death<0.0001,"<0.0001", round(p_val.death, digits = 4))

# p-value of Died in ED (6) : 1 if post_ed=6, 0 otherwise
df.selected$DE <- ifelse(df.selected$post_ed == 6,1,0)
df.selected$DE <- as.factor(df.selected$DE)
test.DE <- chisq_test(DE ~ updated_group, data = df.selected)
p_val.DE <- pvalue(test.DE)
ifelse(p_val.DE<0.0001,"<0.0001", round(p_val.DE, digits = 4))

# p-value of Operating room (2) : 1 if post_ed=2, 0 otherwise
df.selected$OR <- ifelse(df.selected$post_ed == 2,1,0)
df.selected$OR <- as.factor(df.selected$OR)
test.OR <- chisq_test(OR ~ updated_group, data = df.selected)
p_val.OR <- pvalue(test.OR)
ifelse(p_val.OR<0.0001,"<0.0001", round(p_val.OR, digits = 4))

# p-value of Interventional radiology(11) : 1 if post_ed=11, 0 otherwise
df.selected$IR <- ifelse(df.selected$post_ed == 11,1,0)
df.selected$IR <- as.factor(df.selected$IR)
test.IR <- chisq_test(IR ~ updated_group, data = df.selected)
p_val.IR <- pvalue(test.IR)
ifelse(p_val.IR<0.0001,"<0.0001", round(p_val.IR, digits = 4))

# p-value of Intensive care unit (1) : 1 if post_ed=1, 0 otherwise
df.selected$ICU <- ifelse(df.selected$post_ed == 1,1,0)
df.selected$ICU <- as.factor(df.selected$ICU)
test.ICU <- chisq_test(ICU ~ updated_group, data = df.selected)
p_val.ICU <- pvalue(test.ICU)
ifelse(p_val.ICU<0.0001,"<0.0001", round(p_val.ICU, digits = 4))

# p-value of Mortality rate at 24 hours after admission   
df.selected$MR <- ifelse(df.selected$timediff<=24*3600,1,0)
df.selected$MR <- as.factor(df.selected$MR)
test.MR <- chisq_test(MR ~ updated_group, data = df.selected)
p_val.MR <- pvalue(test.MR)
ifelse(p_val.MR<0.0001,"<0.0001", round(p_val.MR, digits = 4))
```

```{r}
num.death <- df.selected%>% filter(death=="Dead")%>%summarise(count = n())
num.death/nrow(df.selected)

num.death24 <- df.selected%>% filter(MR==1)%>%summarise(count = n())
num.death24/num.death
```

## 5. variable analysis in data discription ("dead" vs "not dead") in paper 
```{r}
# percentage of male
df.selected %>% group_by(death) %>% summarise(count = n(), count.M = sum(sex=="Male"), perc = count.M/count)

# test
test.sex <- chisq_test(factor(sex) ~ factor(death), data = df.selected)
p_val.sex <- pvalue(test.sex)
ifelse(p_val.sex<0.0001,"<0.0001", round(p_val.sex, digits = 4))
```

```{r}
# age
df.age1 <- df.selected %>% group_by(death) %>% summarise(age.med = median(age), age.1Q = quantile(age, 0.25), age.3Q = quantile(age, 0.75))
df.age1

# p-value of age
age.test <- kruskal.test(age ~ factor(death),data = df.selected)
p_val.age <- age.test$"p.value" 
ifelse(p_val.age<0.0001,"<0.0001", round(p_val.age, digits = 4))
```

```{r}
# ISS
df.ISS1 <- df.selected %>% group_by(death) %>% summarise(ISS.med = median(injury_serverity_score, na.rm = TRUE), ISS.1Q = quantile(injury_serverity_score, 0.25, na.rm = TRUE), ISS.3Q = quantile(injury_serverity_score, 0.75, na.rm = TRUE))
df.ISS1

# p-value
ISS.test <- kruskal.test(injury_serverity_score ~ death,data = df.selected)
p_val.ISS <- ISS.test$"p.value" 
ifelse(p_val.ISS<0.0001,"<0.0001", round(p_val.ISS, digits = 4))
```


```{r}
# GCS
df.gcs1 <- df.selected %>% 
  group_by(death) %>% 
  summarise(gcs.med = median(gcs_calc_a, na.rm = TRUE), 
            gcs.1Q = quantile(gcs_calc_a, 0.25, na.rm = TRUE), 
            gcs.3Q = quantile(gcs_calc_a, 0.75, na.rm = TRUE))
df.gcs1

# p-value
gcs.test <- kruskal.test(gcs_calc_a ~ updated_group,data = df.selected)
p_val.gcs <- gcs.test$"p.value"
ifelse(p_val.gcs<0.0001,"<0.0001", round(p_val.gcs, digits = 4))
```


```{r}
# heart rate
df.hr0 <- df.selected %>% 
  filter(pulse_a==0) %>% 
  group_by(death) %>% 
  summarise(count_0=n())

df.hr <- df.selected %>% 
  group_by(death) %>% 
  summarise(hr.med = median(pulse_a, na.rm = TRUE), 
            hr.1Q = quantile(pulse_a, 0.25, na.rm = TRUE), 
            hr.3Q = quantile(pulse_a, 0.75, na.rm = TRUE),
            count = n())%>% 
  left_join(df.hr0, by = "death") %>%
  mutate(perc_0 = count_0/count)

df.hr

# p-value
hr.test <- kruskal.test(pulse_a ~ death,data = df.selected)
p_val.hr <- hr.test$"p.value" 
ifelse(p_val.hr<0.0001,"<0.0001", round(p_val.hr, digits = 4))
```

```{r}
# blood pressure
df.bp0 <- df.selected %>% 
  filter(sys_bp_a==0) %>% 
  group_by(death) %>% 
  summarise(count_0=n())

df.bp1 <- df.selected %>% 
  group_by(death) %>% 
  summarise(bp.med = median(sys_bp_a), 
            bp.1Q = quantile(sys_bp_a, 0.25), 
            bp.3Q = quantile(sys_bp_a, 0.75), 
            count = n()) %>% 
  left_join(df.bp0, by = "death") %>%
  mutate(perc_0 = count_0/count)

df.bp1

# p-value
test.bp_zero <- chisq_test(factor(bp_zero) ~ factor(death), data = df.selected)
p_val.bp_zero <- pvalue(test.bp_zero)
ifelse(p_val.bp_zero<0.0001,"<0.0001", round(p_val.bp_zero, digits = 4))
```

```{r}
# injury_type: 0=blunt, 1=penetrating
df.injurytype1 <- df.selected %>% group_by(death) %>% summarise(count = n(), count.penetrating = sum(injury_type), perc = count.penetrating/count)
df.injurytype1

# p-value
test.injurytype <- chisq_test(factor(injury_type) ~ factor(death), data = df.selected)
p_val.injurytype <- pvalue(test.injurytype)
ifelse(p_val.injurytype<0.0001,"<0.0001", round(p_val.injurytype, digits = 4))

```

```{r}
# trauma_center_level: 0=level1, 1=level 2
df.level1 <- df.selected %>% group_by(death) %>% summarise(count = n(), count.level1 = sum(trauma_center_level), perc = count.level1/count)
df.level1

# p-value
test.centerlevel <- chisq_test(factor(trauma_center_level) ~ factor(death), data = df.selected)
p_val.centerlevel <- pvalue(test.centerlevel)
ifelse(p_val.centerlevel<0.0001,"<0.0001", round(p_val.centerlevel, digits = 4))
```

